---
title: "Getting started with R"
author: "Vladan Devedzic"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
### Setup
Download R from <https://cran.r-project.org/> and install it.  
Download RStudio from <https://www.rstudio.com/products/rstudio/download/> and install it.

### Open new project and new R script
Open new project in RStudio (File > New Project...).  
Open new R script in the project (File > New File > R Script).

### Working directory
Get working directory:
```{r eval=FALSE}
getwd()
```
Set working directory:  
Session > Set Working Directory

### Help
```{r eval=FALSE}
?setwd
help("setwd")
```

### Install and load packages
```{r message=FALSE, eval=FALSE}
# install.packages("ggplot2")
library(ggplot2)
# install.packages("caret")
library(caret)
```

## Basic commands, operators and data types

Most of the sections in this document are structured as follows:  

* pseudocode  
* R code examples  
* results or running the R code examples  

In pseudocode throughout this document:  

* syntax like `<something>` is used as a descriptive palceholder for an appropriate, context-specific identifier or another piece of code in lines of an actual R script  
* syntax like `"<something>"` is used as a descriptive palceholder for an appropriate, context-specific *string* in lines of an actual R script  
* leading '+' character in multi-line pseudocommands is used for indentation purposes only; it should be omitted in actual R code

### Print a line of text in the console
```{r}
print("Hi :)")
```

### Naming and coding conventions
See <https://google.github.io/styleguide/Rguide.xml>.

### Assignment statement
`x <- <something>`
```{r}
x <- 2
x
```

### Manipulating objects in the workspace
`ls()                      # list all objects in memory`  
`rm(<o1>, <o2>, <o3>, ...) # remove one or more objects from memory by their names`  
`rm(list = ls())           # remove all objects from memory (usually not recommended)`
```{r eval=FALSE}
ls()
rm(x, y)
```

### Operators
* \+	  Add, 2 + 3 = 5  
* \-	  Subtract, 5 - 2 = 3  
* \*	  Multiply, 2 * 3 = 6  
* \/	  Divide, 6 / 2 = 3  
* \^	  Exponent, 2 ^ 3 = 8  
* \%%	Modulus operator, 9%%2 = 1  
* \%/%	Integer division, 9 %/% 2 = 4  
* \<	  Less than  
* \>	  Greater than  
* \=	  Equal to  
* \<=	Less than or equal to  
* \>=	Greater than or equal to  
* \!=	Not equal to  
* \!	  Not  
* \|	  OR  
* \&	  And

### Expressions
E.g., `x / y - z^2` etc.
```{r}
3.4 / 2 + 7^2
```

### Vectors
`<y> <- c(<something1>, <something2>, <something3>, ...)`  
`<y> <- rep(<something>, <times>)`  
`<y> <- <int1>:<int2>`  
`<y> <- seq(<value1>, <value2>, by = <step>)`  
The index of the first element in a vector is 1, not 0.
```{r}
y <- c(1, 2, 3)
z <- c(1.2, 3)
t <- 2:6
w <- seq(3.2, 4.7, by = 0.2)

w[3]
w[]
w
```

### Matrices
`<m> <- matrix(c(3, 5, 7, 1, 9, 4), nrow = 3, ncol = 2, byrow = TRUE)`  
`<m>.nrow <- nrow(<m>) # number of rows`  
`<m>.ncol <- ncol(<m>) # number of columns`  
`<m> <- t(<m>)         # transpose <m>`  
`<m>[2,3]`  
`<m>[2]`  
`<m>[2, ]`
```{r}
a <- matrix(8:1, nrow = 2, ncol = 4, byrow = TRUE)
a
a.nrow <- nrow(a)
a.nrow
a <- t(a)
a
a[1,2]
a[2, ]
a[2]
a[]
```

### Lists
Ordered collections of elements of different types.  
`<list> <- list(<e1.name> = <e1>, <e2.name> = <e2>, <e3.name> = <e3>, ...)`  
`<list>[[<index>]]     # accessing list element by index, showing value only`  
`<list>[<index>]       # accessing list element by index, showing both name and value`  
`<list>$<element.name> # accessing list element by its name`  
`is.list(<something>)                                  # Is <something> a list?`  
`<combined.list> <- c(<list1>, <list2>, <list3>, ...)  # list concatenation`  
`names(<list>)                           # names of list elements`  
`<list>[names(<list>) == <element.name>] # all elements of a list having the same name`  
`unlist(<list>)                          # convert list into a named character vector`  
`unlist(<list>, use.names = FALSE)       # convert list into a character vector`  
`append(<list>,                          # insert new element into an existing list, after index <n>`  
`+      list(<e1.name> = <e>),           # new element must be a list itself, that's why list(<e1.name> =  <e>)`  
`+           <n>)                        # <n> is optional; if omitted, new element is appended at the end`  
`<list>[[<n>]] <- NULL                   # remove <n>th element from <list>`
```{r}
traveler1 <- list(adult = TRUE, passport = "P212123", age = 34)
traveler1
traveler1[[3]]
traveler2 <- list(adult = FALSE, passport = "P4567756", age = 14)
traveler2
traveler2$age
travelers <- c(traveler1, traveler2)
travelers
travelers[[3]]
travelers[[5]]
travelers[5]
is.list(travelers)
is.vector(travelers)
names(travelers)
travelers[names(travelers) == "age"]
unlist(travelers)
unlist(travelers, use.names = FALSE)
age.of.travelers <- unlist(travelers[names(travelers) == "age"], use.names = FALSE)
age.of.travelers
length(traveler1)
traveler1 <- append(traveler1, list(country = "AUS"), 2)
length(traveler1)
traveler1
traveler1[[3]] <- NULL
length(traveler1)
traveler1
```

### Data types
Vector, factor, numeric, character, logical, data.frame, matrix, list, ...  
`class(<something>)                    # data type`  
`mode(something), typeof(<something>)  # how a data item is internally stored in memory`
```{r}
class(a)
mode(a)
typeof(a)
typeof(2.3)
```

### Factors
`<b> <- c(1, 2, 2, 2, 3, 1, 1, 4, 5, 4)`  
`<f> <- as.factor(b)`  
`levels(<f>)`  
`<f> <- factor(c(1, 2, 3))`  
`<f> <- gl(3,                                                    # gl() "generates levels" (here 3), i.e. factors`  
`+         1, length = 10, labels = c("One", "Two", "Three"))    # each level replicated 1 time, length(<f>) = 10`
```{r}
b <- c(1, 1, 1, 2, 1, 1, 1, 1, 5, 4)
b.as.factor <- as.factor(b)
levels(b.as.factor)
f <- gl(3, 1, length = 10, labels = c("One", "Two", "Three"))
f
f <- gl(3, 2, length = 10, labels = c("One", "Two", "Three"))
f
meal = factor(c("Lunch","Dinner"))
meal
meal = factor(c("Lunch","Dinner"), levels=c("Lunch","Dinner"))
meal
```


### Dataframes
`<dataframe> <- as.data.frame(<matrix>)`  
`str(<dataframe>)`
```{r}
a.data.frame <- as.data.frame(a)
a.data.frame
str(a.data.frame)
```

## Loops and branching
### for, if, break, next
`for (<i> in <int vector>) {`  
`+  <line 1>`  
`+  <line 2>`  
`+  ...`  
`+  if (<logical condition>) {`  
`+    <line i1>`  
`+    <line i2>`  
`+    ...`  
`+    break       # break: exit the loop; next: skip the remaining lines in this iteration`  
`+  }`  
`+  ...`  
`+  <line n>`  
`}`  
```{r}
for (i in 1:10) {
  if (i == 3) {
    print("Done")
    break
  }
  s <- paste(i,"is current index", sep = " ")
  print(s)
}
```

### while, if-else, break, next
`<i> <- <initial value>`  
`while (logical condition involving <i>) {`  
`+  <line 1>`  
`+  <line 2>`  
`+  ...`  
`+  if (<logical condition>) {`  
`+    <line i1>`  
`+    <line i2>`  
`+    ...`  
`+    break       # break: exit the loop; next: skip the remaining lines in this iteration`  
`+  } else {`  
`+    <line j1>`  
`+    <line j2>`  
`+    ...`  
`+  }`  
`+  ...`  
`+  <line n>`  
`+  <i> <- <modify <i>>`  
`}`  
```{r}
i <- 1
while (i <= 10) {
  if (i == 5) {
    i <- i + 1
    next
  } else {
    print(paste(i, "is current index", sep = " "))
    i <- i + 1
  }
}
```

### ifelse(<condition>, v1, v2)
Can return a vector (if <condition> involves another vector).
```{r}
ifelse(1 < 6, TRUE, FALSE)
ifelse(1 < 6, "<", "Not <")
ifelse(1:10 < 6, 1, 2)
```

## ggplot2

```{r}
# install.packages("ggplot2")
library(ggplot2)
```

Data to plot (used in the examples below):
```{r}
actor.xy <- data.frame(year = factor(c(2014, 2015, 2016, 2017)), movies = (c(2, 3, 2, 1)))
actor.xy
```

### Bar graphs
`ggplot(data = <dataframe>, `  
`+      aes(x = <column 1>, y = <column 2>, fill = <column 1>)) +  # fill = <column 1> is optional; no y for counts`  
`+ geom_bar(stat = "identity") +                                   # "identity" for values, "count" for counts`  
`+ xlab("<x-axis label>") + ylab("<y-axis label>") +`  
`+ ggtitle("<graph title>")`  
```{r message=FALSE}
actor.plot.set <- ggplot(data = actor.xy, aes(x = year, y = movies, fill = year))
actor.plot.set + geom_bar(stat = "identity")
actor.plot.set +
  geom_bar(stat = "identity") +
  xlab("recent years") + ylab("# of movies") +
  ggtitle("XY's recent movies")
```

### Line graphs
`ggplot(data = <dataframe>,`  
`+      aes(x = <column 1>, y = <column 2>, group = 1)) +  # group = 1: one line, all points connected`  
`+ geom_line(colour = "<colour>", linetype = "<linetype>", size = <line thickness>) +`  
`+ geom_point(colour="<colour>", size = <point size>, shape = <point shape>, fill = "<point fill colour>") +`  
`+ xlab("<x-axis label>") + ylab("<y-axis label>") +`  
`+ ggtitle("<graph title>")`  
`All parameters in geom_line() and in geom_point() are optional.`  
`The defaults are: colour = "black", linetype = "solid", size = 1, shape = 21 (circle), fill = "black"`  
`See <http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/> for more information on colors.`  
`See <http://www.cookbook-r.com/Graphs/Shapes_and_line_types/> for information on shapes and line types.`

```{r}
actor.plot.set <- ggplot(data = actor.xy, aes(x = year, y = movies, group = 1))
actor.plot.set + geom_line()
actor.plot.set + geom_line() + geom_point()
actor.plot.set + 
  geom_line(colour = "blue", linetype = "dotted", size = 2) + 
  geom_point(colour="green", size = 4, shape = 21, fill = "yellow")
# geom_point(color="green", size = 8, shape = 18, fill = "yellow")
```

## Working with datasets / dataframes

### Reading a dataset
`<dataframe> <- read.csv("<filename>", stringsAsFactors = FALSE)`  
` str(<dataframe>)  # structure of <dataframe>, all variables/columns`  
` head(<dataframe>) # the first few rows`  
` tail(<dataframe>) # the last few rows`
```{r}
the.beatles.songs <- read.csv("The Beatles songs dataset, v1.csv", 
                              stringsAsFactors = FALSE)
```

### Examining a dataframe
`str(<dataframe>)              # structure of <dataframe>, all variables/columns`  
`dim(<dataframe>)              # showing dimensions (numbers of rows and columns) of a dataframe`  
`names(<dataframe>)            # showing column names`  
`head(<dataframe>)             # the first few rows`  
`tail(<dataframe>)             # the last few rows`  
`<dataframe>[ , ]              # the entire dataframe`  
`<dataframe>                   # the entire dataframe`  
`<dataframe>[<m>, ]            # m-th row`  
`<dataframe>[ ,<n>]            # n-th column`  
`summary(<dataframe>$<column>) # summarizing a variable/column values`  
`fix(<dataframe>)              # editing a dataframe`  
`new.df <- edit(<dataframe>)   # editing a dataframe and assigning the modified dataframe to another datavrame`  
```{r}
str(the.beatles.songs)
dim(the.beatles.songs)
names(the.beatles.songs)
head(the.beatles.songs)
tail(the.beatles.songs)
the.beatles.songs[4, ]
the.beatles.songs[ ,2]
summary(the.beatles.songs$Duration)
summary(the.beatles.songs$Title)
summary(the.beatles.songs$Year)
# fix(the.beatles.songs)
# a.data.frame.1 <- edit(a.data.frame)
```

### Examining a dataframe visually, with ggplot()
```{r}
the.beatles.songs.clean <- read.csv("The Beatles songs dataset, v1, no NAs.csv", 
                              stringsAsFactors = FALSE)
the.beatles.songs.clean$Year <- factor(the.beatles.songs.clean$Year)      # because write.csv/read.csv produces int's
g1 <- ggplot(data = the.beatles.songs.clean, aes(x = Year, y = Duration, fill = Year))
g1 + geom_bar(stat = "identity")
g2 <- ggplot(data = the.beatles.songs.clean, aes(x = Year, fill = Year))
g2 + geom_bar(stat = "count") +
  xlab("Year") + ylab("No. of songs") +
  ggtitle("The number Beatles songs per year")
g3 <- ggplot(the.beatles.songs.clean[1:5, ], aes(x = Year, y = Duration, group = 1))
g3 + geom_line(color = "orange", size = 2, linetype = "longdash") + 
     geom_point(color = "red", shape = 25, size = 8, fill = "yellow")
```

### Adding/Removing columns to/from a dataframe
`<dataframe>$<new column name> <- <default value>  # adding a new column (default values)`  
`<dataframe>$<column name> <- NULL                 # removing a column`

```{r}
the.beatles.songs$Not.on.album <- FALSE
the.beatles.songs$Not.on.album <- NULL
the.beatles.songs$On.album <- FALSE
the.beatles.songs$On.album[the.beatles.songs$Album.debut != ""] <- TRUE
```

### Adding new rows to a dataframe
In case of adding one new row, it must be a 1-line dataframe with the same column names. It is also possible to add an entire dataframe to the existing one (with the same column names).  
`<new row> <- data.frame(<column name 1> = <value 1>, <column name 2> = <value 2>,...)`  
`<new data frame> <- rbind(<dataframe>, <new row>) # append new row to the end of the existing dataframe`  
`<new data frame> <- rbind(<dataframe>[1:i, ],     # insert new row in the middle`  
`+                         <new row>,`  
`+                         <dataframe>[(i + 1):nrow(<dataframe>), ])`
```{r}
new.song <- data.frame(the.beatles.songs[1, ])
the.beatles.songs <- rbind(the.beatles.songs, new.song)
the.beatles.songs <- rbind(the.beatles.songs[1:3, ],    # Rstudio keeps the original row numbers in View()
                           new.song, 
                           the.beatles.songs[4:nrow(the.beatles.songs), ])

```

### Removing rows from a dataframe
`<dataframe>[-i, ]                                 # show dataframe without i-th row`  
`<dataframe>[-c(i, j, k), ]                        # show dataframe without rows i, j, k`  
`<dataframe> <- <dataframe>[-i, ]                  # remove i-th row from dataframe`  
`<dataframe> <- <dataframe>[-c(i, j, k), ]         # remove rows i, j, k from dataframe`  
`<dataframe> <- <dataframe>[-(i:k), ]              # remove rows i to k from dataframe`  

```{r}
nrow(the.beatles.songs)
the.beatles.songs <- the.beatles.songs[-nrow(the.beatles.songs), ]
the.beatles.songs1 <- the.beatles.songs[-(305:310), ]
the.beatles.songs <- the.beatles.songs[-(1:304), ]
the.beatles.songs <- rbind(the.beatles.songs1, the.beatles.songs)
```

### Changing column names
`colnames(<dataframe>)[i] <- "<new name>"`  

```{r}
colnames(the.beatles.songs)
which(colnames(the.beatles.songs) == "Genre")
colnames(the.beatles.songs)[which(colnames(the.beatles.songs) == "Genre")] <- "Song.genre"
colnames(the.beatles.songs)[6] <- "Genre"
```

### Changing row names
`rownames(<dataframe>)[i] <- "<new name>"`  
`rownames(<dataframe>) <- c("<new name 1>", "<new name 2>",...)`  
`rownames(<dataframe>) <- c(1, 2,...)`  
`rownames(<dataframe>) <- list("<new name 1>", <numeric 2>,...)`  

```{r}
rownames(the.beatles.songs) <- paste("song", 1:nrow(the.beatles.songs))
rownames(the.beatles.songs) <- c(1:nrow(the.beatles.songs))
```

### Slicing and dicing dataframes
`<selection> <- <dataframe>[<some rows>, <some columns>]`  
`<selection> <- <dataframe>[i:k, c("<column 1>", "<column 2>",...)]`  
`<indexes> <- with(<dataframe>, which(<condition; can be complex>))  # a with()-which() selection, like an SQL query`  
`<selection> <- <dataframe>[<indexes>, ]`  
`<selection> <- subset(<dataframe>,                                  # subset() is much like SELECT... FROM... WHERE`  
`+                     <logical condition for the rows to return>,`  
`+                     <select statement for the columns to return>)   # can be omitted; `  
`+                                                                     # column names not prefixed by <dataframe>$`  
`library(dplyr)`  
`<selection> <- filter(<dataframe>,                                    # filter() is from dplyr`  
`+                     <logical condition for the rows to return>)     # can include column referencing, `  
`+                                                                     # not-prefixed by <dataframe>$`
```{r}
selected.songs <- the.beatles.songs[1:5, c("Title", "Album.debut")]
# View(selected.songs)
indexes <- with(the.beatles.songs, which((Year == "1964") & (Lead.vocal != "McCartney")))
selected.songs <- the.beatles.songs[indexes, ]
songs.1958 <- subset(the.beatles.songs, Year == 1958, c("Title", "Album.debut"))
library(dplyr)
filter(the.beatles.songs, 
       as.integer(rownames(the.beatles.songs)) < 33 & Title == "12-Bar Original")

```

### Shuffling rows/columns
`<dataframe> <- <dataframe>[sample(nrow(<dataframe>)), ]   # shuffle row-wise`  
`<dataframe> <- <dataframe>[, sample(ncol(<dataframe>))]   # shuffle column-wise`  

```{r}
the.beatles.songs <- the.beatles.songs[sample(nrow(the.beatles.songs)), ]
the.beatles.songs <- the.beatles.songs[, sample(ncol(the.beatles.songs))]
```

### Replacing selected values in a column
`<selected var name> <- <dataframe>$<column> == <selected value>`  
`<dataframe>$<column>[<selected var name>] <- <new value>`  

```{r}
empty.album.debut <- the.beatles.songs$Album.debut == ""
empty.album.debut
the.beatles.songs$Album.debut[empty.album.debut] <- "empty"
the.beatles.songs$Album.debut[empty.album.debut] <- ""
```

### Applying functions to all elements in rows/columns of a dataframe

`apply(<dataframe>, <1 | 2>, <function(x) {...}>)  # 1 | 2: apply function(x) by row | column`  
`mapply(function(x, y, ...) {...}, <dataframe>$<column 1>, <dataframe>$<column 2>, ...)`  
`+  # <dataframe>$<column 1> corresponds to x, <dataframe>$<column 2> corresponds to y, ...`  
`+  # alternatively:  <f> <- function(x, y, ...) {...}`  
`+                    mapply(<f>, <dataframe>$<column 1>, <dataframe>$<column 2>, ...)`  
`+                      # <f> is just the function name (!)`  
`+                      # <dataframe>$<column 1> corresponds to x, <dataframe>$<column 2> corresponds to y, ...,`  
`+                      # or can be columns from different dataframes, "independent" vectors,... (of the same length)`  
`sapply(<vector>, FUN = function(x) {...})   # function(x): function to be applied to each element of <vector>`

```{r}
apply(the.beatles.songs[1, ], 1, function(x) {print(x)})
apply(the.beatles.songs[1, ], 2, function(x) {print(x)})
mapply(function(x, y) {print(x); print(y)}, 
       the.beatles.songs[111:113, ]$Title, 
       the.beatles.songs[111:113, ]$Year)
sapply(the.beatles.songs[1, ], FUN = function(x) {print(x)})
```

### Partitioning a dataframe
`# install.packages('caret')`  
`library(caret)`  
`set.seed(<any specific int>)  # allows for repeating the randomization process exactly`  
`<indexes> <- createDataPartition(<dataframe>$<column>, p = 0.8, list = FALSE)`  
`<partition 1> <- <dataframe>[<indexes>, ]`  
`<partition 2> <- <dataframe>[-<indexes>, ]`  

```{r message=FALSE}
library(caret)
set.seed(222)
indexes <- createDataPartition(the.beatles.songs$Year, p = 0.8, list = FALSE)
the.beatles.songs.p1 <- the.beatles.songs[indexes, ]
the.beatles.songs.p2 <- the.beatles.songs[-indexes, ]
```

### Saving a dataset (modified or newly created dataset)
`write.csv(x = <dataframe>, file = "<filename>", row.names = F)  # do not include the row names (row numbers) column`  
`saveRDS(object = <dataframe or another R object>, file = "<filename>")  # save R object for the next session`  
`<dataframe or another R object> <- readRDS(file = "<filename>")         # restore R object in the next session`

```{r}
write.csv(the.beatles.songs.p2, "p2.csv", row.names = F)
saveRDS(the.beatles.songs.p2, "p2.RData")
p2 <- readRDS("p2.RData")
```

## Data type conversion
`# Covered above:`  
`# b <- c(1, 2, 2, 2, 3, 1, 1, 4, 5, 4)`  
`# b.as.factor <- as.factor(b)`  
`# levels(b.as.factor)`  
`# e.g., <dataframe> <- as.data.frame(<matrix>)`  
`# str(<dataframe>)`  

### Difference between character and factor vectors
`summary(<character vector>)`  
`summary(as.factor(<character vector>))`  
```{r}
class(the.beatles.songs$Year)
summary(the.beatles.songs$Year)
summary(as.factor(the.beatles.songs$Year))
```

### Convert numeric to factor
`<dataframe>$<numeric column with few different values> <-`  
`+ factor(<dataframe>$<numeric column with few different values>,`  
`+        levels = c(0, 1, ..., k), labels = c("<l1>", "<l2>", ..., "<lk>"))`
```{r}
the.beatles.songs1 <- the.beatles.songs
the.beatles.songs1$Billboard.hit <- 0
the.beatles.songs1$Billboard.hit[!is.na(the.beatles.songs1$Top.50.Billboard)] <- 1
the.beatles.songs1$Billboard.hit <- 
  factor(the.beatles.songs1$Billboard.hit, levels = c(0,1), labels = c("N", "Y"))
class(the.beatles.songs1$Billboard.hit)
summary(the.beatles.songs1$Billboard.hit)
levels(the.beatles.songs1$Billboard.hit)
```

### Examples
#### Fixing some values in the.beatles.songs$Year
```{r}
summary(the.beatles.songs$Year)
summary(as.factor(the.beatles.songs$Year))
the.beatles.songs$Year[the.beatles.songs$Year == "196?"] <- "1969"
the.beatles.songs$Year[the.beatles.songs$Year == "1977/1994"] <- "1977"
the.beatles.songs$Year[the.beatles.songs$Year == "1980/1995"] <- "1980"
summary(as.factor(the.beatles.songs$Year))
```

#### Creating the.beatles.songs$Billboard.hit column as factor
```{r}
the.beatles.songs$Billboard.hit <- 0
the.beatles.songs$Billboard.hit[!is.na(the.beatles.songs$Top.50.Billboard)] <- 1
the.beatles.songs$Billboard.hit <- 
  factor(the.beatles.songs$Billboard.hit, levels = c(0,1), labels = c("N", "Y"))
```

## Working with tables

### The table() function
`table(<var>)  # typically a factor or an integer var`  
```{r}
table(the.beatles.songs1$Year)
table(the.beatles.songs1$Top.50.Billboard)
table(the.beatles.songs1$Billboard.hit)
table(the.beatles.songs1$Billboard.hit)[1]
x <- table(the.beatles.songs1$Billboard.hit)[1]
x
y <- as.numeric(x)
y
```

### The prop.table() function
`prop.table(table(<var>))`  
`round(prop.table(table(<var>)), digits = <n>)`  
```{r}
prop.table(table(the.beatles.songs1$Billboard.hit))
round(prop.table(table(the.beatles.songs1$Billboard.hit)), digits = 2)
```
### Row and column margins
`table(<var1>, <var2>)                                   # <var1>, <var2>: usually factors or integers`  
`table(<rows title> = <var1>, <columns title> = <var2>)  # add common titles for rows/columns`  
`prop.table(table(<var1>, <var2>), margin = 1)           # all row margins are 1.0`  
`prop.table(table(<var1>, <var2>), margin = 2)           # all column margins are 1.0`  

```{r}
table(the.beatles.songs$Billboard.hit, the.beatles.songs$Year)
table(Hit = the.beatles.songs$Billboard.hit, Year = the.beatles.songs$Year)
round(prop.table(table(the.beatles.songs$Billboard.hit, the.beatles.songs$Year), 1), digits = 2)
round(prop.table(table(the.beatles.songs$Billboard.hit, the.beatles.songs$Year), 2), digits = 2)
round(prop.table(table(the.beatles.songs$Billboard.hit, the.beatles.songs$Year)), digits = 2)
```

### Example: converting the.beatles.songs$Year to factor and showing it in tables
```{r}
factor(the.beatles.songs$Year)
the.beatles.songs$Year <- as.factor(the.beatles.songs$Year)
class(the.beatles.songs$Year)
summary(the.beatles.songs$Year)
prop.table((table(the.beatles.songs$Year)))
round(prop.table((table(the.beatles.songs$Year))), digits = 2)
```

### The xtabs() function
`xtabs(~<column 1> + <column 2>, <dataframe>)`  
```{r}
xtabs(~Billboard.hit + Year, the.beatles.songs)
```

## Working with vectors

### Differences in initializing vectors and dataframe columns
`<vector> <- rep(<value>, <times>)`  
`<vector> <- <value>`  
`<dataframe>$<column> <- rep(<value>, <times>)`  
`<dataframe>$<column> <- <value>`  
```{r}
v <- rep(0, 5)
v
v <- 2
v
df <- data.frame(a = c(1, 2, 3), b = c(4, 5, 6))
df
df$a <- rep(1, 3)
df
df$a <- 0
df
```

### Counting the number of elements with the values of `<x>` in a vector
1.  `<table> <- table(<vector>)`  
`  <table>`   
`  <table>[names(<table>) == <x>]`  
2.  `sum(<vector> == <x>)`
3.  `length(which(<vector> == <x>))    # which() is like WHERE in SQL`
```{r}
v <- c(1, 2, 1, 3, 2, 4, 5, 1, 3, 1)
t <- table(v)
t
t[names(t) == 1]
t[names(t) == "1"]
sum(v == 1)
length(which(v == 1))
```

### Appending an element to a vector
`<vector> <- append(<vector>, <element>)   # type conversion occurs if <element> is of different type than v[i]`  
`<vector> <- append(<vector>, <element>, after = <n>)   # insert <=> append at a desired location`  
`<vector> <- append(<vector>, NA)`  
```{r}
v <- c(1, 2, 1, 3, 2, 4, 5, 1, 3, 1)
v
v <- append(v, NA)
v <- append(v, NA, after = 5)
v
v <- append(v, "s")
v
```

### Removing NAs from a vector in NA-sensitive functions
`<function>(<vector>, na.rm = TRUE)`  
```{r}
v <- c(1, 2, 1, 3, 2, 4, 5, 1, 3, 1)
v
v <- append(v, NA)
v <- append(v, NA)
mean(v)
mean(v, na.rm = TRUE)
```

### Selecting vector elements that match criteria, with controlling for NAs and NaNs

`<numeric vector> <- c(<n1>, <n2>, <n3>, ..., NA, ...NaN)`  
`<selected> <- <numeric vector>[<logical criterion> & !is.na(<numeric vector>)]  # is.na() is TRUE for both NA and NaN`  
Using `is.na()` is the only way to test if `<something>` is NA (`<something> == NA` does not work).
```{r}
v <- c(1, 2, 1, 3, NA, 4, 5, 1, 3, NaN, 1)
v
v <- v[v > 1 & !is.na(v)]
v
```

## Working with strings

### Some of the basic string functions
`# install.packages("stringr")`  
`library(stringr)`  
`nchar(<s>)                              # string length`  
`str_length(<s>)                         # string length; str_length() is from stringr`  
`substr(<s>, <start index>, <end index>) # substring`  
`toupper(<s>)                            # to upper case letters`  
`tolower(<s>)                            # to lower case letters`  
`grepl(<s1>, <s2>)                       # contains; TRUE if <s2> contains <s1>`  
`str_detect(<s1>, <s2>)                  # contains; TRUE if <s1> contains <s2>; str_detect() is from stringr`  
`paste(<s1>, <s2>, sep = "")             # concatenate (result: <s1><s2>; <s1> <s2>, if sep = "" omitted)`  
`sub(<s1>, <s2>, <s>)                    # substring replacement: replace <s1> in <s> with <s2>`  
`strsplit(<s>, <regex>)                  # split (the type of the result is list)`  
```{r}
# install.packages("stringr")
library(stringr)
title <- the.beatles.songs$Title[13]
title
nchar(title)
str_length(title)
grepl("You", title)
str_detect(title, "You")
```

### Splitting strings to words
`strsplit(<s>, <regex>)                  # split (the type of the result is list)`  
```{r}
title <- the.beatles.songs$Title[13]
words.in.title <- strsplit(title, " ")
words.in.title
words.in.title <- strsplit(title, " ")
words.in.title
words.in.title <- strsplit("All  My    Loving", " ")
words.in.title
words.in.title <- unlist(words.in.title)
words.in.title <- words.in.title[words.in.title != ""]
words.in.title
title <- paste(words.in.title[1], words.in.title[2], words.in.title[3])
title
title <- paste(words.in.title[1], words.in.title[2], words.in.title[3], sep = "")
title
```

## Resources, readings, references

R Tutorials, <http://www.endmemo.com/program/R/>  
R: A Beginner's Guide (by Sharon Machlis), <http://www.tfrec.wsu.edu/TFREConly/r4beginners_v3.pdf>  
Graphs with ggplot2, <http://www.cookbook-r.com/Graphs/>
